Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2023-12-25 paper.tex
I'm tired

The paper is organized as follows: in \cref{sec:methods} we provide
mathematical background and describe methods for computation of 3-point
correlation functions. Verification of these methods along with some results for
real-world samples (limestone, soil etc.) is described in
\cref{sec:results}. Section \ref{sec:discussion} provides the discussion, a
comparison with other existing methods and summarizes possible directions for
improvement.

@@

\subsection{Analysis of anisotropic media}
When an analized sample is anisotropic, i.e. it is stretched or oriented along
some specific direction, it may be desired to align one of the catheti of the
scanning pattern \cref{fig:pattern} with that direction. In this case the
anisotropicity present in the sample will be featured in a produced correlation
function \textcolor{red}{most well. voobshe eto po-angliyski?}

To detect the direction of anisotropy we use an $N \times N$ (where $N$ is
dimensionality of the sample) covariance matrix for coordinates of points in the
phase of interest:
\begin{equation}
  C_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and decompose $C$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation (by
multiplication of one of the columns by $-1$ if needed, so that
$\det T = 1$. Application of this rotation to the sample makes the direction of
anisotropicity parallel to one of the axes, so anisotropicity can be better
featured in the correlation function \cref{fig:aniso}.
\begin{figure*}[tp]
  \centering
  \subfigure[An original image of an anisotropic pore.]{
    \includegraphics[width=0.4\linewidth, frame]{images/aniso.png}
    \label{fig:aniso-unrot}}
  \hfill
  \subfigure[A rotated image of the pore, so it is elongated along the axis
    $X$.]{
    \includegraphics[width=0.4\linewidth, frame]{images/aniso-rot.png}
    \label{fig:aniso-rot}}
  \vskip\baselineskip
  \subfigure[$S_3$ function for the original image]{
    \includegraphics[width=0.3\linewidth]{images/aniso-s3.png}}
  \hfill
  \subfigure[$S_3$ function for the rotated image. A vector $(0.92, -0.4)$ is a
    direction of elongation of the pore.]{
  \includegraphics[width=0.3\linewidth]{images/aniso-rot-s3.png}}
  \hfill
  \subfigure[$S_3$ function averaged across 100 random rotations.]{
    \includegraphics[width=0.3\linewidth]{images/aniso-avg-s3.png}
    \label{fig:aniso-avg}}
  \caption[]{Behaviour of our algorithm for calculation of the $S_3$ function
    when dealing with highly anisotropic media. The pattern \cref{fig:pattern}
    which we use in our algorithm is most sensitive to anisotropy which is
    directed along one of the axes. A result on \cref{fig:aniso-avg} averaged
    across rotations is similar to the algorithm proposed by Malmir~et~al.}
  \label{fig:aniso}
\end{figure*}
The above said holds for every algorithm introduced in the paper.

@@

\subsection{Applications to material images}
In this section we compute correlation functions along the plane $XY$ for 2- and
3-dimensional real samples. Table \ref{tab:summary} contains information about
these samples. Note that we compute surface correlation functions only for those
samples for which the criterion $C_{0.5} > 0.97$ holds (the sandstone and the
limestone). Images of the samples are on \cref{fig:samples}. On plots of
correlation functions we provide a family of curves with the first argument of
the correlation function being fixed. For two highly anisotropic samples (the
soil and the shale) we provide heatmaps of $S_3$ correlation function where
anisotropy can be easily observed. These two samples were preliminary rotated by
$\frac{\pi}{4}$ around the axis $Z$ before calculation of correlation functions.

Correlation functions for the XCT sample of the limestone are on
\cref{fig:sample-limestone-cf}. The plot of $S_3$ function allows us to estimate
porosity of the sample to be $\sim 0.04$ (note, that the curve with $r_1 = 0$ is
actually $S_2$ function). The two-point correlation function reaches plateau at
$r_2 \approx 200\div 225\ \mu m$ which means that two events of finding two
points in the void phase are independent if the points are farther than
$225\ \mu m$ from each other. We see that the sample is slightly anisotropic
($S_3(0, 225) \approx S_3(150, 0)$). From analysis of $C_3$ function it can be
stated that the sample has clusters of void phase which are $\sim 200\ \mu m$ in
diameter.

Correlation functions for the SEM sample (the sandstone) are on
\cref{fig:sample-sandstone-cf}. Porosity of the sample is $\sim 0.3$ as follows
from the plot of $S_3$ function at $r_1 = 0$. The funcion $S_3(0, r_2)$ reaches
plateau at $r_2 \approx 20\ \mu m$. The sample is highly anisotropic (elongated
along the ordinate). The samples has clusters which are $\sim 45\ \mu m$ in
diameter.

@@

\begin{table*}[!htp]
  \centering
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    Sample name & Image type & Dimensions (pixels) & Image resolution ($\mu m$) & $C_{0.5}$\\
    \hline
    Limestone & XCT &  $600 \times 600 \times 600$ & 3.02 & 0.9712 \\
    Soil & XCT & $160 \times 160 \times 160$ & 2.25 & 0.9210 \\
    Shale & XCT & $160 \times 160 \times 160$ & 2.25 & 0.8912 \\
    Sandstone & SEM &  $1280 \times 869$ & 0.10 & 0.9703 \\
    \hline
  \end{tabular}
  \caption{Summary of samples used for calculation of correlation functions.}
  \label{tab:summary}
\end{table*}

\begin{figure*}[tp]
  \centering
  \subfigure[Limestone]{
    \includegraphics[width=0.4\linewidth]{images/carb.png}
    \label{fig:sample-limestone}}
  \hfill
  \subfigure[Sandstone (different clusters of void phase are shown in color)]{
    \includegraphics[width=0.4\linewidth, frame]{images/sandstone1.png}
    \label{fig:sample-sandstone}}
  \vskip\baselineskip
  \subfigure[Soil]{
    \includegraphics[width=0.4\linewidth]{images/soil.png}
    \label{fig:sample-soil}}
  \hfill
  \subfigure[Shale]{
    \includegraphics[width=0.4\linewidth]{images/shale.png}
    \label{fig:sample-shale}}
  \caption[]{Images of samples used for calculation of higher-order correlation
    functions.}
  \label{fig:samples}
\end{figure*}

\begin{figure*}[tp]
  \centering
  \includegraphics[width=0.4\linewidth]{images/carb.npy-surf3.png}
  \hfill
  \includegraphics[width=0.4\linewidth]{images/carb.npy-surf2void.png}
  \vskip\baselineskip
  \includegraphics[width=0.4\linewidth]{images/carb.npy-surfvoid2.png}
  \hfill
  \includegraphics[width=0.4\linewidth]{images/carb.npy-s3.png}
  \vskip\baselineskip
  \includegraphics[width=0.4\linewidth]{images/carb.npy-c3.png}
  \caption[]{Correlation functions for carbonate sample
    (\cref{fig:sample-limestone}).}
  \label{fig:sample-limestone-cf}
\end{figure*}

\begin{figure*}[tp]
  \centering
  \includegraphics[width=0.4\linewidth]{images/sandstone1.png-surf2void.png}
  \hfill
  \includegraphics[width=0.4\linewidth]{images/sandstone1.png-surfvoid2.png}
  \vskip\baselineskip
  \includegraphics[width=0.4\linewidth]{images/sandstone1.png-s3.png}
  \hfill
  \includegraphics[width=0.4\linewidth]{images/sandstone1.png-c3.png}
  \caption[]{Correlation functions for sandstone sample
    (\cref{fig:sample-sandstone}).}
  \label{fig:sample-sandstone-cf}
\end{figure*}

@@

\textcolor{red}{Where to put this?}
We have made a comparison of our algorithm with two algorithms, one of which
proposed by Malmir~et~al \cite{malmir2018} and another proposed by Smith and
Torquato \cite{SMITH1988176}. The first algorithms generates a triangle with
fixed sides $r_1$ and $r_2$ and an angle $\theta$ between them and places it in
the medium firstly making a random translation and a rotation around a random
vector by a random angle. By making a large number of trials the algorithm gets
a value of a three-point correlation function at the point
$(r_1, r_2, \theta)$. The second algorithm takes a precalculated triangle
$(r_1, r_2, \theta)$ from a web-like pattern and places it in the medium without
rotation, only applying a random translation. This way the algorithm preserves
orientation of triangles stored in the pattern.

@@

Our algorithm fixes $\theta = \frac{pi}{2}$ and does not perform random
rotations or trial triangles. It is sensitive to anisotropy of the medium
when it is stretched along one of the axes as shown in \cref{fig:aniso}. A
direction of elongation for a $N$-dimensional image can be easily detected by
firstly evaluating a $N \times N$ covariance matrix for points belonging to the
phase of interest:
\begin{equation}
  C_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and then decomposing $C$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation and
can be used to align the elongation with the axes (compare \cref{fig:aniso-unrot}
and \cref{fig:aniso-rot}).

The main advantage of our algorithm is its parallelizability and ease of
implementation. An equivalent operation for rotation of test triangles is a
rotation of the whole input image which we have implemented in our
CorrelationFunctions.jl package. By applying a rotation to the input first we
made a rest of the algorithm cache-friendly by avoiding random memory access.

>>>>

The paper is organized as follows: in \cref{sec:methods} we provide
mathematical background and describe methods for computation of 3-point
correlation functions. Verification of these methods along with some results for
real-world samples (sandstone, soil etc.) is described in
\cref{sec:results}. Section \ref{sec:discussion} provides the discussion, a
comparison with other existing methods and summarizes possible directions for
improvement.

@@

\subsection{Analysis of anisotropic media}
When an analized sample is anisotropic, i.e. it is stretched or oriented along
some specific direction, it may be desired to align one of the catheti of the
scanning pattern \cref{fig:pattern} with that direction. In this case the
anisotropy present in the sample will be featured in a produced correlation
function \textcolor{red}{most well. voobshe eto po-angliyski?}

To detect the direction of anisotropy we use an $N \times N$ (where $N$ is
dimensionality of the sample) covariance matrix for coordinates of points in the
phase of interest:
\begin{equation}
  C_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and decompose $C$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation (by
multiplication of one of the columns by $-1$ if needed, so that
$\det T = 1$. Application of this rotation to the sample makes the direction of
anisotropy parallel to one of the axes, so anisotropy can be better featured in
the correlation function \cref{fig:aniso}.
\begin{figure*}[tp]
  \centering
  \subfigure[An original image of an anisotropic pore.]{
    \includegraphics[width=0.4\linewidth, frame]{images/aniso.png}
    \label{fig:aniso-unrot}}
  \hfill
  \subfigure[A rotated image of the pore, so it is elongated along the axis
    $X$.]{
    \includegraphics[width=0.4\linewidth, frame]{images/aniso-rot.png}
    \label{fig:aniso-rot}}
  \vskip\baselineskip
  \subfigure[$S_3$ function for the original image]{
    \includegraphics[width=0.3\linewidth]{images/aniso-s3.png}}
  \hfill
  \subfigure[$S_3$ function for the rotated image. A vector $(0.92, -0.4)$ is a
    direction of elongation of the pore.]{
    \includegraphics[width=0.3\linewidth]{images/aniso-rot-s3.png}
    \label{fig:aniso-rot-s3}}
  \hfill
  \subfigure[$S_3$ function averaged across 100 random rotations.]{
    \includegraphics[width=0.3\linewidth]{images/aniso-avg-s3.png}
    \label{fig:aniso-avg}}
  \caption[]{Behaviour of our algorithm for calculation of the $S_3$ function
    when dealing with highly anisotropic media. The pattern \cref{fig:pattern}
    which we use in our algorithm is most sensitive to anisotropy which is
    directed along one of the axes. A result on \cref{fig:aniso-avg} averaged
    across rotations is similar to the algorithm proposed by Malmir~et~al.}
  \label{fig:aniso}
\end{figure*}
The above said holds for every algorithm introduced in the paper.

@@

\subsection{Applications to material images}
In this section we compute correlation functions for 2- and 3-dimensional real
samples. Table \ref{tab:summary} contains information about these samples. Note
that we compute surface correlation functions only for those samples for which
the criterion $C_{0.5} > 0.97$ holds (the two sandstones and the \textcolor{red}{shale}). Images
of the samples are on \cref{fig:samples}. On plots of correlation functions for
isotropic samples we provide a family of curves with the first argument of the
correlation function being fixed. For anisotropic samples we provide heatmaps of
correlation functions where anisotropy can be easily observed.

Correlation functions for the XCT sample of a sandstone are on
\cref{fig:sample-sandstone-xct-cf}. The plot of $S_3$ function allows us to
estimate porosity of the sample to be $\sim 0.19$ (note that the curve with
$r_1 = 0$ is actually $S_2$ function). The two-point correlation function
reaches plateau at $r_2 \approx 60 \mu m$ which means that two events
of finding two points in the void phase are independent if the points are
farther than $60 \mu m$ from each other. We see that the sample is isotropic
($S_3(0, 56.25) \approx S_3(56.25, 0)$). From similarity of the $C_3$ and $S_3$
functions it can be stated that the pore phase in this sample forms one big
interconnected pore system.

Correlation functions for the SEM sample (the sandstone) are on
\cref{fig:sample-sandstone-cf}. Porosity of the sample is $\sim 0.3$ as follows
from the plot of $S_3$ function at $r_1 = 0$. The function $S_3(0, r_2)$ reaches
plateau at $r_2 \approx 20\ \mu m$. The sample is highly anisotropic (elongated
along the ordinate). The samples has clusters which are $\sim 45\ \mu m$ in
diameter.

@@

\begin{table*}[!htp]
  \centering
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    Sample name & Image type & Dimensions (pixels) & Image resolution ($\mu m$) & $C_{0.5}$\\
    \hline
    Sandstone & XCT &  $500 \times 500 \times 500$ & 2.25 & 0.9703 \\
    Soil & XCT & $160 \times 160 \times 160$ & 2.25 & 0.9210 \\
    Shale & XCT & $160 \times 160 \times 160$ & 2.25 & 0.8912 \\
    Sandstone & SEM &  $1280 \times 869$ & 0.10 & 0.9703 \\
    \hline
  \end{tabular}
  \caption{Summary of samples used for calculation of correlation functions.}
  \label{tab:summary}
\end{table*}

\begin{figure*}[tp]
  \centering
  \subfigure[Sanstone XCT]{
    \includegraphics[width=0.4\linewidth]{images/15581.png}
    \label{fig:sample-sandstone-xct}}
  \hfill
  \subfigure[Sandstone SEM (different clusters of void phase are shown in color)]{
    \includegraphics[width=0.4\linewidth, frame]{images/sandstone1.png}
    \label{fig:sample-sandstone}}
  \vskip\baselineskip
  \subfigure[Soil]{
    \includegraphics[width=0.4\linewidth]{images/soil.png}
    \label{fig:sample-soil}}
  \hfill
  \subfigure[Shale]{
    \includegraphics[width=0.4\linewidth]{images/shale.png}
    \label{fig:sample-shale}}
  \caption[]{Images of samples used for calculation of higher-order correlation
    functions.}
  \label{fig:samples}
\end{figure*}

\begin{figure*}[tp]
  \centering
  \includegraphics[width=0.4\linewidth]{images/15581-surf3.png}
  \hfill
  \includegraphics[width=0.4\linewidth]{images/15581-surf2void.png}
  \vskip\baselineskip
  \includegraphics[width=0.4\linewidth]{images/15581-surfvoid2.png}
  \hfill
  \includegraphics[width=0.4\linewidth]{images/15581-s3.png}
  \vskip\baselineskip
  \includegraphics[width=0.4\linewidth]{images/15581-c3.png}
  \caption[]{Correlation functions for the sandstone XCT sample
    (\cref{fig:sample-sandstone-xct}).}
  \label{fig:sample-sandstone-xct-cf}
\end{figure*}

\begin{figure*}[tp]
  \centering
  \includegraphics[width=0.4\linewidth]{images/sandstone1.png-surf2void.png}
  \hfill
  \includegraphics[width=0.4\linewidth]{images/sandstone1.png-surfvoid2.png}
  \vskip\baselineskip
  \includegraphics[width=0.4\linewidth]{images/sandstone1.png-s3.png}
  \hfill
  \includegraphics[width=0.4\linewidth]{images/sandstone1.png-c3.png}
  \caption[]{Correlation functions for the sandstone SEM sample
    (\cref{fig:sample-sandstone}).}
  \label{fig:sample-sandstone-cf}
\end{figure*}

@@

\textcolor{red}{Where to put this?}
We have made a comparison of our algorithm with two algorithms, one of which
proposed by Malmir~et~al \cite{malmir2018} and another proposed by Smith and
Torquato \cite{SMITH1988176}. The first algorithms generates a triangle with
fixed sides $r_1$ and $r_2$ and an arbitrary angle $\theta$ between them and
places it in the medium firstly making a random translation and a rotation
around a random vector by a random angle. By making a large number of trials the
algorithm gets a value of a three-point correlation function at the point
$(r_1, r_2, \theta)$. The second algorithm takes a precalculated triangle
$(r_1, r_2, \theta)$ from a web-like pattern and places it in the medium without
rotation, only applying a random translation. This way the algorithm preserves
orientation of triangles stored in the pattern. The main advantage of our
algorithm is its parallelizability and ease of implementation. A drawback is
that an angle between catheti of the sampling pattern is fixed to be
$\theta = \frac{\pi}{2}$ but the pattern can be rotated as a whole by an
equivalent operation of rotation of a sample. These rotations can be random as
in the Malmir's algorithm (\cref{fig:aniso-avg}), or it can be a single rotation
to align catheti of the sampling triangle with the direction of anisotropy
(\cref{fig:aniso-rot-s3}).

@@
